Gap formation and stability in non-isothermal protoplanetary discs

### 2015-03-27 paper/results.tex
revision 3

{\bf For completeness we also simulated a locally
  isothermal disc with $h=0.05$ where the
  sound-speed is kept strictly equal to its initial value.  This
  simulation yield a most unstable growth rate
  $\sigma/\Omega(r_p)\simeq 0.05$ at $m=5$,   
  compared with a value of $0.085$ at $m=6$ for a corresponding
  simulation that includes the energy equation but with rapid cooling
  $\tilde{\beta}=0.01$. However, the vortex evolution is similar.}

@@

Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
varies with the cooling rate: for 
$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
$t\sim800P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
$t\sim1200P_0$. This non-monotonic dependence suggests that there
exists an optimal cooling rate to maximise the vortex lifetime. We
will discuss this issue further in 
\S\ref{lifetime_discuss}. Notice the decay timescale decreases with
increasing cooling time: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$
whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
amplitude to decay significantly after reaching maximum.

@@

In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
, but becomes more compact approaching a ratio of 2 during its decay ($t=1310$).   
Notice in Fig. \ref{shockplot} the appearance of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes.

@@

We now examine vortex lifetimes as function of the imposed cooling
times. {\bf For this study, additional simulations with 
  $\tilde\beta=0.01,\,0.25,\,0.75,\,2.5,\,7.5$ were also performed.} 
We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximized for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
$\tilde{\beta}=0.5$. This is due to the longer decay timescale with
decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximizes at $\tilde\beta=2.5$ with 
$t_{\mathrm{life}}\approx 1650P_0$.  The other timescales $t_\mathrm{diss}$ and
$t_\mathrm{Mach}$ are less complicated, but still display a
non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
$\tilde{\beta}=1$.

{\bf We comment here that a locally isothermal simulation, where the
  disc sound-speed is kept constant in time, was also
  performed for comparison. In this case we did not observe
  significant vortex decay within the simulation timescale
 (implying $t_\mathrm{diss}\gtrsim2000P_0$).  We expect the corresponding vortex
 lifetime to exceed that for $\tilde{\beta}=0.01$. Including an energy
 equation with rapid cooling (or setting $\gamma$ close to 
 unity), could still lead to discrepancies with a locally isothermal disc. This is 
 due to the advection-creation of specific entropy within the planet's horseshoe
 region with the former case,  thereby affecting the generalized vortensity 
 and therefore the instability (\S\ref{linear}) and subsequent vortex evolution. 
 Nevertheless, a longer vortex 
 lifetime in a locally isothermal disc would 
 be consistent with the above trend of increasing vortex lifetimes as
 $\tilde{\beta}\to0$.}

@@

In the previous section, we observed that vortices began to decay when
it starts to induce shocks. We therefore suggest the vortex lifetime
is determined the time needed for the vortex to grow to 
sufficient amplitude to induce shocks in the surrounding fluid, and 
the decay timescale itself. The latter was found to shorten 
with increasing $\tilde{\beta}$. However, the former timescale
($t_\mathrm{diss}$), which may be considered as the 
duration of the quasi-steady state, reaches a single maximum with
respect to cooling time. The sum of these two timescales results in the
double-peaked behaviour of $t_\mathrm{life}$ in Fig. \ref{betaplot}. 
We suggest below competing factors that may result in this
non-monotonic dependence of $t_\mathrm{diss}$ on the cooling rate.

>>>>

 For completeness we also simulated a locally
  isothermal disc with $h=0.05$ where the
  sound-speed is kept strictly equal to its initial value.  This
  simulation yield a most unstable growth rate
  $\sigma/\Omega(r_p)\simeq 0.05$ at $m=5$,   
  compared with a value of $0.085$ at $m=6$ for a corresponding
  simulation that includes the energy equation but with rapid cooling
  $\tilde{\beta}=0.01$. However, the vortex evolution is similar.

@@

Fig. \ref{lifetimeplot} shows that the duration of the quasi-steady state
varies with the cooling rate: for 
$\tilde{\beta}=0.1$ and $10$, the vortex amplitude begins to decay around
$t\sim800P_0$, while for $\tilde{\beta}=1,\,5$ the decay begins at 
$t\sim1200P_0$. This non-monotonic dependence suggests that there
exists an optimal cooling rate to maximise the vortex lifetime. We
will discuss this issue further in 
\S\ref{lifetime_discuss}. Notice the decay timescale {\bf can be long} with
rapid cooling: for $\tilde{\beta}=0.1$ it takes $\sim400P_0$ 
whereas for $\tilde{\beta}=10$ it takes $\sim 100P_0$ for the $m=1$
amplitude to decay significantly after reaching maximum.

@@

In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
, but becomes more compact approaching a ratio of 2 during its decay ($t=1310P_0$).   
Notice in Fig. \ref{shockplot} the appearance of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes.

@@

We now examine vortex lifetimes as function of the imposed cooling
times. For this study, additional simulations with 
  $\tilde\beta=0.01,\,0.25,\,0.75,\,2.5,\,7.5$ were also performed.  
We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximized for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
$\tilde{\beta}=0.5$. {\bf Note that for very small $\tilde{\beta}$, 
  there is significant contribution to the overall vortex lifetime due
  to a long decay timescale. }
For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximizes at $\tilde\beta=2.5$ with 
$t_{\mathrm{life}}\approx 1650P_0$.  
%The other timescales $t_\mathrm{diss}$ and
%$t_\mathrm{Mach}$ are less complicated, but still display a
%non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
%$\tilde{\beta}=1$.

 We comment here that a locally isothermal simulation, where the
  disc sound-speed is kept constant in time, was also
  performed for comparison. In this case we did not observe
  significant vortex decay within the simulation timescale
 (implying $t_\mathrm{diss}\gtrsim2000P_0$).  We expect the corresponding vortex
 lifetime to exceed that for $\tilde{\beta}=0.01$. Including an energy
 equation with rapid cooling (or setting $\gamma$ close to 
 unity), could still lead to discrepancies with a locally isothermal disc. This is 
 due to the advection-creation of specific entropy within the planet's horseshoe
 region with the former case,  thereby affecting the generalized vortensity 
 and therefore the instability (\S\ref{linear}) and subsequent vortex evolution. 
 Nevertheless, a longer vortex 
 lifetime in a locally isothermal disc would 
 be consistent with the above trend of increasing vortex lifetimes as
 $\tilde{\beta}\to0$.

@@

In the previous section, we observed that vortices began to decay when it starts to
induce shocks. {\bf We thus suggest that the time needed for the vortex to grow to 
sufficient amplitude to induce shocks in the surrounding fluid,   
%The former timescale
%($t_\mathrm{diss}$), 
which may be considered as the duration of the quasi-steady state or
$t_\mathrm{diss}$, is an important contribution to the overall vortex
lifetime.} 
%reaches a single maximum with
%respect to cooling time {\bf for $\tilde{\beta}\geq 0.1$ (see also Fig. \ref{lifetimeplot})}. 
%{\bf (The case with $\tilde{\beta}=0.01$ and the locally isothermal disc, 
%does not follow this broad trend and may need to be analyzed separately.)} 
%The sum of these two timescales leads to the 
%double-peaked behaviour of $t_\mathrm{life}$ in Fig. \ref{betaplot}. 
We discuss below some competing factors that may result in a 
non-monotonic dependence of $t_\mathrm{diss}$ on the cooling rate.

### 2015-03-27 paper/summary.tex
updating
revision 3

The duration of the quasi-steady state increases with
increasing cooling timescales until a critical value, beyond which this
quasi-steady state shortens again. The timescale for the vortex to
decay, after reaching maximum amplitude, decreases monotonically with
increasing cooling time. We do observe vortex migration during this
  decay, which may influence this decay timescale.  
The sum of
these two timescales results in a double-peaked vortex lifetime as a
function of cooling timescale. 
%technically we should probably say there is an optimum cooling time
%to maximise the quasi-steady state

%Our results indicate there exists an optimum cooling timescale to
%maximise the lifetime of gap-edge vortices.

We suggest the non-monotonic dependence \emph{of the quasi-steady state} can
be attributed to the time required for the vortex to grow to 
sufficient amplitude to induce shocks in the surrounding fluid,
thereby losing energy and also smooth out the gap edge.

For short cooling timescales, the planet is able to open a
deeper gap which favours the RWI, leading to stronger 
vortices. For long cooling timescales, we find the vortex
grows faster during the quasi-steady state. In accordance with
previous stability calculations \citep{li00}, we 
suggest the latter is due to the RWI being favoured with increasing
disc temperature, and that this effect overcomes weaker
gap-opening for sufficiently long cooling times. 
These competing factors imply
for both very short and long cooling timescales, the vortex reaches
its maximum amplitude, shock, and begins to decay, sooner than
intermediate cooling timescales. 
%{\bf, for which the quasi-steady state is 
%the longest.}

%decay timescale depends on cooling time, but there seems to be a
%monotonic trend: longer cooling times, decay timescale is shorter, so
%this  
%doesn't contribute to non-monotonic trend

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simulations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, $h$ is a dynamical
variable, but by analyzing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
the longest vortex lifetime, that  
$h\approx0.058$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \cite{fu14}.

@@

  \emph{Cooling model.} Our prescription for the disc 
  heating/cooling is convenient to probe the full range of
  thermodynamic response of the disc. However, in order to calculate 
  vortex lifetimes in actual 
  protoplanetary discs, an improved thermodynamics treatment ,
  e.g. radiative cooling based on realistic disc temperature, density,
  opacity models etc., should be used in future work.

>>>>

{\bf For our main simulations with $\tilde{\beta}\geq 0.1$}, the duration of the quasi-steady state increases with
increasing cooling timescales until a critical value, beyond which this
quasi-steady state shortens again. {\bf We find the timescale for the vortex to
decay, after reaching maximum amplitude, can be long for small
$\tilde{\beta}$ which contributes to a long overall vortex lifetime
with rapid cooling.} We do observe vortex migration during {\bf its}
decay, which may influence this decay timescale.

% The sum of
% these two timescales results in a double-peaked vortex lifetime as a
% function of cooling timescale.

We suggest a non-monotonic dependence of the quasi-steady state {\bf
on the cooling timescale $\tilde{\beta}$} can
be attributed to the time required for the vortex to grow to 
sufficient amplitude to induce shocks in the surrounding fluid,
thereby losing energy and also smooth out the gap edge.

For short cooling timescales, the planet is able to open a
deeper gap which favours the RWI, leading to stronger 
vortices. For long cooling timescales, we find the vortex
grows faster during the quasi-steady state. In accordance with
previous stability calculations \citep{li00}, we 
suggest the latter is due to the RWI being favoured with increasing
disc temperature, and that this effect overcomes weaker
gap-opening for sufficiently long cooling times. 
These competing factors imply
for both short and long cooling timescales, the vortex reaches
its maximum amplitude, shock, and begins to decay, sooner than
intermediate cooling timescales. {(\bf However, for very rapid
  cooling, e.g. $\tilde{\beta}=0.01$, the quasi-steady state is also
  quite long. In this case other factors may be important, such as
  vortex migration.)}

%{\bf, for which the quasi-steady state is 
%the longest.}      
%decay timescale depends on cooling time, but there seems to be a
%monotonic trend: longer cooling times, decay timescale is shorter, so
%this  
%doesn't contribute to non-monotonic trend

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simulations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, $h$ is a dynamical
variable, but by analyzing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
the longest vortex lifetime {\bf in the presence of moderate cooling}, that  
$h\approx0.058$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is consistent with \cite{fu14}.

@@

  \emph{Cooling model.} Our prescription for the disc 
  heating/cooling is convenient to probe the full range of
  thermodynamic response of the disc. However, in order to calculate 
  vortex lifetimes in actual 
  protoplanetary discs, an improved thermodynamics treatment,
  e.g. radiative cooling based on realistic disc temperature, density,
  opacity models etc., should be used in future work.
